Universal algebra and the \texorpdfstring{$ \lambda $}{lambda}-calculus

### 2024-08-24 report/src/category-theory.tex
Show that the Yoneda embedding preserves binary products and exponential objects

Now, this embedding has a couple of properties:
\begin{lemma}
  For any $ c: C $ and $ F : PC $, we have an equivalence $ PC(Y(c), F) \simeq F(c) $, and this equivalence is natural in $ c $ and $ F $.
\end{lemma}
\begin{proof}
  See \autocite{Kashiwara}, Proposition 1.4.3.
\end{proof}

\begin{lemma}
  The Yoneda embedding functor preserves limits.
\end{lemma}
\begin{proof}
  See \autocite{borceux}, Volume 1, Proposition 2.15.5.
\end{proof}

\begin{lemma}
  The Yoneda embedding functor is \iindex{cartesian}. That is, not only does it preserve binary products, but it also preserves exponential objects (\autocite{stackexchange:yoneda-exponentials}).
\end{lemma}
\begin{proof}
  \TODO
\end{proof}

>>>>

Now, this embedding has a couple of properties:
\begin{lemma}
  For any $ A: C $ and $ F : P C $, we have an equivalence $ P C(Y(A), F) \simeq F(A) $, and this equivalence is natural in $ c $ and $ F $.
\end{lemma}
\begin{proof}
  It sends a natural transformation $ \alpha: P C(Y(A), F) $ to the element $ \alpha_A(\id A) $. Conversely, it sends an element $ x : P(A) $ to the natural transformation $ \alpha $ given by $ \alpha_B(f) = P(f)(x) $ for all $ f: Y(A)(B) = C(B, A) $. For more details, see \autocite{Kashiwara}, Proposition 1.4.3.
\end{proof}

\begin{remark}
  For any category $ D $, the category $ [C, D] $ has (co)limits of some kind (terminal or initial object, (co)products, (co)equalizers) iff $ D $ does. These (co)limits are computed pointwise: for example, for binary products, $ (F \times G)(X) = F(X) \times G(X) $. In particular $ P C $ has all limits and colimits because $ \SET $ has all limits and colimits.
\end{remark}

\begin{remark}
  Also, for any small category $ C $ (small means that the collection of objects is a type in our type universe. For example: $ \TYPE $ or $ \SET $ are not small categories, because a type cannot contain itself), the presheaf category $ P C $ has exponential objects, given by
  \[ (F^G)(X) = \hom{PC}(Y(X) \times G, F) \]
  the natural transformations from the product functor of the Yoneda embedding of $ X $ and $ G $, to $ F $ (see \autocite{MacLane-Moerdijk}, Section I.6, Proposition 1).
\end{remark}

\begin{definition}
  Suppose that we have a functor $ F: C \to D $, and $ C $ and $ D $ both have binary products. Then for $ A, B : C $, we have morphisms $ \pi_1: A \times B \to A $ and $ \pi_2 : A \times B \to B $ and applying $ F $ to these yields morphisms from $ F(A \times B) $ to $ F(A) $ and $ F(B) $. We then have a product morphism
  \[ \langle F(\pi_1), F(\pi_2) \rangle: F(A \times B) \to F(A) \times F(B). \]
  Now, if this morphism is an isomorphism for all $ A, B : C $, we say that $ F $ \index{preservation of binary products}\textit{preserves binary products}.

  One can imagine that there exists a similar definition of preservation of limits in general, in which the limit morphism of $ F $ applied to the projections from the limit is an isomorphism for any limit in $ C $.
\end{definition}

\begin{definition}\label{def:exponentials-preservation}
  Suppose that we have a functor $ F: C \to D $ that preserves binary products, and $ C $ and $ D $ have exponential objects. Then for $ A, B : C $, we have natural bijections
  \[ \varphi: C(A^B \times B, A) \xrightarrow \sim C(A^B, A^B) \quad \text{and} \quad \psi: D(F(A^B) \times F(B), F(A)) \xrightarrow \sim D(F(A^B), F(A)^{F(B)}) \]
  This gives a morphism $ F(\varphi^{-1}(\id{A^B})) : D(F(A^B \times B), F(A)) $. Precomposing with the inverse of the isomorphism $ f: F(A^B \times B) \xrightarrow \sim F(A^B) \times F(B) $ and applying $ \psi $ gives
  \[ \psi(f^{-1} \cdot \varphi^{-1}(\id{A^B})) : D(F(A^B), F(A)^{F(B)}). \]
  We say that $ F $ \index{preservation of exponentials}\textit{preserves exponentials} if this morphism is an isomorphism for all $ A, B : C $. A name for a functor that preserves exponentials is a\iindex{cartesian functor}.
\end{definition}

\begin{lemma}
  The Yoneda embedding preserves limits.
\end{lemma}
\begin{proof}
  See \autocite{borceux}, Volume 1, Proposition 2.15.5 for the full proof, but note that for binary products, we have bijections
  \[ Y(A \times B)(X) \xrightarrow \sim Y(A)(X) \times Y(B)(X) \]
  sending $ f : C(X, A \times B) $ to the pair $ (f \cdot \pi_1, f \cdot \pi_2) $ and conversely, sending a pair $ (g_1, g_2) : C(X, A) \times Y(X, B) $ to the product morphism $ \langle g_1, g_2 \rangle : Y(X, A \times B) $. This idea is the core of the proof about all limits.
\end{proof}

\begin{lemma}
  The Yoneda embedding preserves exponentials.
\end{lemma}
\begin{proof}
  First of all, note that for $ A, B, X : C $, we have a sequence of isomorphisms  (\autocite{stackexchange:yoneda-exponentials})
  \begin{align*}
    Y(A^B)(X) &\cong P C(Y(X), Y(A^B))\\
    &\cong C(X, A^B)\\
    &\cong C(X \times B, A)\\
    &\cong P C(Y(X \times B), Y(A))\\
    &\cong P C(Y(X) \times Y(B), Y(A))\\
    &\cong P C(Y(X), Y(A)^Y(B))\\
    &\cong (Y(A)^Y(B))(X)
  \end{align*}
  using (once or twice) the Yoneda lemma, fully faithfulness of the Yoneda embedding, the property of the exponential object, and the fact that the Yoneda embedding preserves binary products.

  Some calculating shows that when applying this isomorphism to some $ f: Y(A^B)(X) $, we get the natural transformation $ h: P C(Y(X) \times Y(B), Y(A)) $ given by
  \[ h_Z(g_1, g_2) = \langle g_1, g_2 \rangle \cdot \varphi^{-1}(f) \]
  for all $ Z: C $ and $ (g_1, g_2) : Y(X)(Z) \times Y(B)(Z) $ and with the natural bijection
  \[ \varphi: C(X \times B, A) \xrightarrow \sim C(X, A^B). \]

  It turns out that when we apply the morphism in Definition \ref{def:exponentials-preservation} to $ f $, we get exactly the same natural transformation. Therefore, the morphism in Definition \ref{def:exponentials-preservation} is the isomorphism defined here and we conclude that $ Y $ preserves exponentials.
\end{proof}
